Research Spike: Evaluation of Using a Single Mnemonic both for Cryptocurrency and Identity Wallets

### 2021-12-22 prism-backend/research/single-mnemonic/bip32-etal.tex
Added comments and refinements based on yesterday's sync.

>>>>

\paragraph{Notation.} It is frequent that the values used to derive keys in
the different layers are written in a short form like \texttt{m/a'/b'/c'/d'/e/f},
where \texttt{m} denotes the master seed (mnemonic); and each value between
bars is the number used to derive the key for the node at the corresponding
layer, where a tilded value denotes hardened derivation, and a non-tilded value
denotes non-hardened derivation. For instance, in the previous path, \texttt{a'}
denotes that the value \texttt{a} is used to derive, in hardened mode, the second
level of the path; or the value \texttt{f}  is used to derive, in non-hardened
mode, the last node in the path.

### 2021-12-22 prism-backend/research/single-mnemonic/discussion.tex
Added comments and refinements based on yesterday's sync.

>>>>

From a theoretical point of view, it seems very reasonable to derive all
keys, even for different purposes (payments, staking, or identities) from
the same mnemonic. \secref{ssec:unification}, listed a series of topics
that should be taken into account prior to unifying Cardano and Atala
wallets under the same mnemonic, to avoid security pitfalls. In addition to
those,  there are also some aspects that may need careful consideration, as they
may have impact in the overall security and privacy properties of the wallet
(and related systems).

\paragraph{Malleability considerations when combining multiple
  attributes into one address.} %
Are the keys for the different purposes going to be used \emph{only} in a
standalone manner? Or, rather, it can be expected that they are somehow
encoded and distributed jointly? The latter seems to be the case for PoS
wallets, which are expected to be used in the Cardano ecosystem. As pointed
out by \cite{kkl20}, naive encoding of keys with different purposes, into
a single address, may enable certain (malleability) attacks.

\paragraph{Concrete differences in prior security analysis.} %
The security analysis in \cite{def+21} (which is one of the main references
used in this research spike) is on BIP32, not BIP44. While BIP44
is actually a subset of BIP32, it further specifies it in concrete ways
that may alter the result of the analysis. For instance, \cite{def+21}
assumes that non-hardened nodes are leaves of the tree (which, alternatively,
can be seen as roots in new trees). However, in BIP44 the leaves are
non-hardened nodes, and intermediate nodes are mostly hardened. Yet another
possible source of discrepancy is that the analysis assumes that non-hardened
nodes are maintained in a hot/cold wallet setting and, thus, the private keys
cannot be compromised. This assumption may not hold for all use cases, and
that would change completely the security reasoning. If we want to determine
the concrete security level we want/need in some specific use cases that differ
from the analysis of \cite{def+21} as mentioned above, a new model might be
necessary.

\paragraph{Careful consideration of extra security properties.} %
The main related work analysed so far\footnote{BIP32, BIP44, \cite{def+21}
  and, partially, \cite{kkl20}.} study the security of either BIP32
wallets for payments, or PoS wallets. The introduction of identity-related
(frequently associated to actual people) data may require additional properties,
only mentioned in these works (or not considered, as in the BIPs). For instance,
some sort of forward security will be very desirable. While, for the case of
payments or staking, the impact of a compromise (either of a leaf node --
i.e., a key -- or a complete subtree) can be mitigated by prompt detection
and transfer of funds to an uncompromised address, this may be harder for
keys representing in some way real world identities. Most probably, the
latter will frequently be long-lived keys, that also can, in turn, be used
to issue further identities (e.g., as in the case of identity providers).
Thus, the consequences of a compromise may be more cumbersome to address
(re-issuing credentials, distributing revocation lists, etc.) and,
consequently, it seems desirable to look for constructions that give some
sort of guarantee about the security of keys produced at time $t' < t$, if a
compromise happens at time $t$.

Also, it cannot be discarded that further desirable privacy or security
properties arise, as we get more familiar with the targetted use cases.

\subsection{Further related work}

There seems to be a growing body of related research, which is specially
growing lately (probably, due to the rise of cryptocurrencies). The following
just includes some of the main references for self-bookkeeping (besides
\cite{kkl20} and \cite{def+21}, already mentioned in earlier sections). Note
that they, in turn, include references to further related work.

\begin{itemize}
\item ``Arcula: A Secure Hierarchical Deterministic Wallet for Multi-asset
  Blockchains'', \cite{lfa20}.
\item ``Simple, Efficient and Strongly KI-Secure Hierarchical Key Assignment
  Schemes'', \cite{fpp13}.
\item ``A Formal Treatment of Deterministic Wallets'', \cite{dfl19}.
\end{itemize}

### 2021-12-22 prism-backend/research/single-mnemonic/iowal.tex
Added comments and refinements based on yesterday's sync.

\paragraph{Lightwallet.} HD wallets in Cardano are mostly BIP44-compatible
wallets, with the exception that the used elliptic curve is Ed25519,
instead of Secp256k1, like Bitcoin's BIP32 (and thus, BIP44). This
difference is marked by using index 1852 for hardened child derivation at
layer 1 (purpose level), as specified in CIP1852%
\footnote{\url{https://cips.cardano.org/cips/cip1852/}. Last access, December
  15th, 2021.}.

\paragraph{Atala.} HD wallets in Atala follow a slightly different tree
organisation, layer-wise. Namely, it defines 4 layers: root, DID number, key
type, and index. All child derivations are done in hardened mode. The elliptic
curve is Secp256k1, as BIP44. See the key derivation document for details%
\footnote{\url{https://github.com/input-output-hk/atala-prism/blob/master/prism-backend/docs/protocol/key-derivation.md}. Last access December 15th, 2021. \textbf{WARNING!
    Internal link!}}.

@@

>>>>

\paragraph{Lightwallet.} HD wallets in Cardano are mostly BIP44-compatible
wallets, with the exception that the used elliptic curve is Ed25519,
instead of Secp256k1, like Bitcoin's BIP32 (and thus, BIP44). This
difference is marked by using index 1852 for hardened child derivation at
layer 1 (purpose level), as specified in CIP1852%
\footnote{\url{https://cips.cardano.org/cips/cip1852/}. Last access, December
  15th, 2021.}. Mnemonics in Cardano are composed by 24 words, and hence
encode 256-bit random bitstrings.

\paragraph{Atala.} HD wallets in Atala follow the same overall rules as in
BIP32. Specifically, the same distinction between hardened and non-hardened
child derivation rules are applied, as well as Secp256k1 as the elliptic curve,
wich ECDSA as digital signature scheme. However, Atala defines a different
tree structure, with the following layers, depicted also in \figref{fig:atalatree}:

\begin{description}
\item[Layer 0: Root.] Encoded as/derived from a 12-word mnemonic.
\item[Layer 1: DID Number.] Obtained by hardened child derivation from the root.
  Each node in this level corresponds to a different DID.
\item[Layer 2: Key Type.]  Obtained by hardened child derivation from a layer 1
  node. Currently, the Atala specification defines four types of keys: master
  keys, issuing keys, communication keys, and authentication keys.
\item[Layer 3: Index.] Obtained by hardened child derivation from a layer 2 node.
  Each layer 2 node may have up to $2^{31}$ child nodes, which are the leaves of
  the tree.
\end{description}

See the key derivation document for details%
\footnote{\url{https://github.com/input-output-hk/atala-prism/blob/master/prism-backend/docs/protocol/key-derivation.md}. Last access December 15th, 2021. \textbf{WARNING!
    Internal link!}}. Mnemonics in Atala are composed by 12 words, and hence
encode 256-bit random bitstrings.

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figures/atala_tree.png}
  \caption{Atala tree derivation structure.}
  \label{fig:atalatree}
\end{figure}

@@

\subsection{Aspects to consider before unification}
\label{ssec:unification}

Next, we emphasize some concrete topics that would need to be considered
from an implementation/software-design point of view, prior to making
Atala and Cardano keys to be derived from a single mnemonic.

\paragraph{Domain separation when different cryptosystems is
  necessary.} %
As mentioned, Cardano uses EdDSA (which is based on Ed25519 curve),
while Atala uses ECDSA with Secp256k1. It is \emph{not} a good idea%
\footnote{I have not been able to find a paper that describes
  some concrete related attack or gives some impossibility result for
  proving security under this circumstance but, at the least, it seems to
  be folklore knowledge. See Lindell's answer at
  \url{https://crypto.stackexchange.com/a/54666/52362} for instance.
  \cite{dlp12+,thorm21} seem good references to study this topic further, if
  needed.} to use the same key in different cryptosystems, nor for different
curves, even though ``structurally'' it may be possible (e.g., in EdDSA and
ECDSA, private keys are 32-byte random numbers). However, this is easily
avoidable by ensuring domain separation in the key derivation functions that
are applied on the common seed. This is an important concern on its own; yet,
it should be addressed natively through the considerations in the next
paragraph.

\paragraph{Ensure a correct hierarchical key derivation strategy.} %
We need to take into account what specific usage we expect from the wallets
and, from there, define hierarchical derivation rules that ensure security
without disabling current utility. Related to the previous paragraph, we should
make sure that keys that are aimed to be used in different
cryptosystems, are derived in separate tree branches that ensure domain
separation. Or, since the Atala derivation path is shorter, than no Atala
derivation path can be a prefix of a Cardano derivation path. For instance,
according to the current specification, the derivation path
\texttt{m/1852'/1815'/0'} corresponds to account \texttt{0'} for Ada coins in
CIP1852-compliant Cardano wallets. In Atala, \texttt{1852'} could be a valid
DID number and, while \texttt{1815'} is currently not a valid key type, if,
in the future it becomes so, then \texttt{m/1852'/1815'/0'} would be a valid
derivation path in both Atala and Cardano -- which could open an attack vector.
An easy fix would be to concatenate the master seed with some value \texttt{C}
in Cardano, and some distinct value \texttt{A} in Atala. The derivation paths would
become \texttt{Cm'/1852'/1815'/...} in Cardano, and \texttt{Am'/...} in Atala
-- hence, domain separation would be ensured. Probably, many other alternatives
exist, more suitable from an engineering perspective (e.g., assigning Atala its
own \texttt{purpose} level in BIP44 wallets). But, whatever option is followed,
domain separation should be ensured.

\paragraph{On the mnemonic length.} %
Atala uses 12-word mnemonics to encode the master seed from which all keys are
derived, while Cardano uses 24 words. According to BIP39, 12 words encode 128-bits
seeds, and 24 words encode 256-bit seeds. These entropy bits are then subject to
the extract-and-expand approach of the HKDF used in BIP32 to produce pseudoranom
bits that are computationally indistinguishable from true random numbers%
\footnote{Note: only 256 of these bits are used to produce the final keys.}.
The keys used by both Atala (for ECDSA), and Cardano (for EdDSA), are thus
securely derived pseudorandom keys of 256 bits. For keys of 256 bits, both ECDSA
and EdDSA are expected to provide 128 bits of security against unforgeability
attacks. Given this, and assuming (like \cite{def+21} does) that some keys
will be compromised, there are two ways to attack a wallet: by breaking
security of the HKDF used to derive the keys; or by breaking security of
the wallet construction itself. Given this, on the one hand, in \cite{def+21}%
\footnote{Important! Note that \cite{def+21} is defined for ECDSA-based BIP32
  wallets with a structure slightly different from BIP44 wallets. Hence, its
  results may not translate to our setting -- and of course, neither to
  EdDSA-based BIP44 wallets.}, security of BIP32 wallets against unforgeability
is estimated to have a security loss of 37 bits over the underlying digital
signature scheme (under somewhat arbitrary estimations for
a typical setting, like leaking about $1\%$ of roughly $2^{20}$ produced
keys). For 256-bit ECDSA keys, this translates to 91 bits of security.
On the other hand, even for 128-bit seeds (thus, with entropy at most
128), breaking the security of the HMAC-based HKDF construction (that would
seem to affect the unlinkability property of BIP32 wallets, rather than their
unforgeability) would be harder than breaking the unforgeability property of
the BIP32 wallet, as per the results in \cite{kraw10}\footnote{According to the
  analysis in \cite{kraw10}, the outputs of HMAC-based HKDFs are roughly
  $q2^{-m}$-close to uniform, where $q$ in our setting would be $2^{20}$ as per
  \cite{def+21}, and $m$ is the min-entropy of the randomness source (our
  mnemonic). Hence, for seeds (mnemonics) of 128 bits of entropy, we can still
  expect the outputs to be $\sim 2^{-100}$-close to uniform.}. Given this, and
lacking a more detailed analysis, seeds of 12 words would appear to be enough
for security, strictly speaking. However, taking into account that we are
aiming to combine two wallets into one, it does not seem logical to reduce the
overall security of the wallet that uses 24-word mnemonics, as the unified
wallet will use the same mnemonic to produce even more keys than what it
was producing before. Therefore, it seems advisable to increase the mnemonic
length of Atala to 24 words, rather than reducing the length of Cardano to
12 words.
